Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-03-01 analysis/paper/paper.Rmd
Changes in Methods and Introduction UP TO the "interaction" section.
saving here to make it easier to revert if unhappy with new intro focusing on hab diff with no frag x xlimate interaction).

<!-- Underemphasize the interaction between fragmentation and drought bc can't do the interaction in test. Instead it's "compare drought effects in fragmented and cf populations."  -->
<!-- In response to this I deleted #1; not looking at interaciton of multiple effects and not studying population dyanmics per se in this one. Suggest alluding to this in Discussion instead. -->

 <!-- [1]. Although the local extinction of species from fragments is common [1], the demographic mechanisms underlying these extinctions are rarely known [2]. However, it is often assumed – especially in the tropics – that these extinctions are driven by the dramatically altered abiotic conditions in fragments [3] reducing rates of reproduction or survivorship [4-6]. This is why understanding the way in which landscape structure and environmental conditions interact to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [7,8].  -->

@@

Populations in fragmented tropical landscapes are also at risk from climate change *(CITATIONS_5)*. These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] as well as increased drought frequency and severity [@caiIncreasingFrequencyExtreme2014] *(add CITATIONS_6)*.
Severe droughts in the Amazon have been shown to reduce aboveground biomass through increased tree mortality and dieback [@williamsonAmazonianTreeMortality2000;[@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008; @lewis2010AmazonDrought2011].

4.  It has been hypothesized that drought and fragmentation could act in synergy to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].
    Fragmentation drives several changes in microclimate near forest edges, including increased air temperature, decreased relative humidity, reduced soil moisture, and increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].
    This decrease in microclimatic buffering could exacerbate the effects of drought in fragments compared to intact forest.
    This may be why the drought-induced mortality of trees is higher near forest edges than in forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
    Alternatively, plant populations persisting in fragments may be more resilient to drought if individuals susceptible to desiccation died quickly after fragment isolation [@betts2019].

5.  The effects of drought and fragmentation could differ among demographic vital rates such as growth, survival, and reproduction.
    In addition to variation among vital rates, responses to drought and fragmentation could depend on life history stage or plant size.
    Large established adults could respond differently than seedlings.
    This is unknown because there are very few systems where we have data on entire life history in a climate change or fragmentation context.
    <!--# I'm not actually testing a size x climate or size x fragmentation interaction explicitly.  Maybe I should not emphasize the variation among size classes in the introduction -->

    -   Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]). Smaller/larger plants might be more susceptible to drought. So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.

@@

>>>>

@@

Populations in fragmented tropical landscapes are also at risk from climate change *(CITATIONS_5)*. These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] as well as increased drought frequency and severity [@caiIncreasingFrequencyExtreme2014] *(add CITATIONS_6)*.
Severe droughts in the Amazon have been shown to reduce aboveground biomass through increased tree mortality and dieback [@williamsonAmazonianTreeMortality2000;[@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008; @lewis2010AmazonDrought2011]. These changes in forest structure can further exascerbate changes in microclimate near the edges of forest fragments, including increased air temperature, decreased relative humidity, reduced soil moisture, and increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].

It has been hypothesized that drought and fragmentation could act in synergy to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999].

    This decrease in microclimatic buffering could exacerbate the effects of drought in fragments compared to intact forest.
    This may be why the drought-induced mortality of trees is higher near forest edges than in forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].
    Alternatively, plant populations persisting in fragments may be more resilient to drought if individuals susceptible to desiccation died quickly after fragment isolation [@betts2019].

The effects of drought and fragmentation could differ among demographic vital rates such as growth, survival, and reproduction.

@@

Future directions:

* There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartz2019]). Smaller/larger plants might be more susceptible to drought. So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.

* Synergistic effects of fragmentation and climate (need the stats to catch up with us)
